Towards Patient-Specific Neural Field Modeling using Micro-Electrode Data

### 2011-04-27 ltx/data_paper/DataPaper.tex
minor edits to EM paer
added second derivative test for maximisation step in data paper

Note that we ignore $\ln\mid\tilde{\boldsymbol\Sigma}_e\mid $ as it is not a function of the parameter set.

@@

Consider first the maximisation of $\mathcal{Q}$-function with respect to $\boldsymbol\theta$, we can write
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=\frac{2}{\sigma_d^2}\boldsymbol\Xi_3-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top(\boldsymbol\Xi_4^\top+\boldsymbol\Xi_4)-\frac{2\xi}{\sigma_d^2}\boldsymbol\Xi_5,
\end{equation}
variable $\boldsymbol\Xi_4$ is symmetric and positive-definite by construction, equating \eqref{eq:Qpartialtheta} to the zero vector yields
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\boldsymbol\Xi_4^{-1}\left(\boldsymbol\Xi_3^\top-\xi\boldsymbol\Xi_5^\top \right).
\end{equation}
The second derivative is negative definite at this stationary point, and hence it is a global maximizer of $\mathcal{Q}$-function. In order to maximise $\mathcal{Q}$-function with respect to $\xi$ we can write
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial\xi}=\frac{2}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -\frac{2\xi}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\frac{2}{\sigma_d^2}\boldsymbol\Xi_5,
\end{equation}
setting \eqref{eq:Qpartialxi} to zero and solving for $\xi$
\begin{equation}
 \xi=\frac{\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\boldsymbol\Xi_5\boldsymbol\Xi_4^{-1}\boldsymbol\Xi_3^\top}{\mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\boldsymbol\Xi_5\boldsymbol\Xi_4^{-1}\boldsymbol\Xi_5^{\top}}
\end{equation}

Now maximizing with respect to $\sigma_{\epsilon}^2$ we get
\begin{equation}
  \frac{\partial \mathcal Q}{\partial \sigma_{\epsilon}^2}=-(T-1)n_y\frac{1}{\sigma_{\epsilon}^2}+\frac{1}{\sigma_{\epsilon}^4}\mathrm{tr}\left\lbrace\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace
\end{equation}
setting above to zero and solving for $\sigma_{\epsilon}^2$ gives
\begin{equation}
 \sigma_{\epsilon}^2=\frac{1}{(T-1)n_y}\sum_{t=0}^{T-1}\mathrm{tr}\left\lbrace (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top \right\rbrace
\end{equation}
A second derivative test shows that this is indeed the maximum. Finally differentiating with respec to $\sigma_d^2$ we get
\begin{align}
 \frac{\partial \mathcal Q}{\partial \sigma_d^2}&= -(T-1)n_x\frac{1}{\sigma_d^2}+\frac{1}{\sigma_d^4}\mathrm{tr}\left\lbrace \boldsymbol\Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\frac{2\xi}{\sigma_d^4} \mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{\xi^2}{\sigma_d^4}\mathrm{tr} \left\lbrace\boldsymbol\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace  \nonumber \\
&-\frac{2}{\sigma_d^4}\boldsymbol\Xi_3\boldsymbol\theta +\frac{1}{\sigma_d^4}\boldsymbol\theta^\top \boldsymbol\Xi_4\boldsymbol\theta+\frac{2\xi}{\sigma_d^4} \boldsymbol\Xi_5 \boldsymbol\theta 
\end{align}
setting above to zero and solving for $\sigma_e^2$ we have
\begin{align}
\sigma_d^2=\frac{1}{(T-1)n_x}\left[ \mathrm{tr}\left\lbrace \boldsymbol\Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-
2\xi\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\xi^2\mathrm{tr} \left\lbrace\boldsymbol\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -2\boldsymbol\Xi_3\boldsymbol\theta+\boldsymbol\theta^\top\boldsymbol \Xi_4\boldsymbol\theta+2\xi\boldsymbol\Xi_5 \boldsymbol\theta\right]  
\end{align}
The second derivative test confirms this is a maximum.

>>>>

@@

Consider first the maximisation of $\mathcal{Q}$-function with respect to $\xi$, we can write
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial\xi}=\frac{2}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -\frac{2\xi}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\frac{2}{\sigma_d^2}\boldsymbol\Xi_5,
setting \eqref{eq:Qpartialxi} to zero and solving for $\xi$ gives
 \xi=\frac{\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\boldsymbol\Xi_5\boldsymbol\Xi_4^{-1}\boldsymbol\Xi_3^\top}{\mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\boldsymbol\Xi_5\boldsymbol\Xi_4^{-1}\boldsymbol\Xi_5^{\top}}
In order to maximise $\mathcal{Q}$-function with respect to $\boldsymbol\theta$ we can write
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=\frac{2}{\sigma_d^2}\boldsymbol\Xi_3-\frac{1}{\sigma_d^2}\boldsymbol\theta^\top(\boldsymbol\Xi_4^\top+\boldsymbol\Xi_4)-\frac{2\xi}{\sigma_d^2}\boldsymbol\Xi_5,
variable $\boldsymbol\Xi_4$ is symmetric and positive-definite by construction, equating \eqref{eq:Qpartialtheta} to the zero vector yields
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\boldsymbol\Xi_4^{-1}\left(\boldsymbol\Xi_3^\top-\xi\boldsymbol\Xi_5^\top \right).
The second derivative is negative definite at this stationary point, and hence it is a global maximizer of $\mathcal{Q}$-function. Now maximizing with respect to $\sigma_{\epsilon}^2$ we get
  \frac{\partial \mathcal Q}{\partial \sigma_{\epsilon}^2}=-(T-1)n_y\frac{1}{\sigma_{\epsilon}^2}+\frac{1}{\sigma_{\epsilon}^4}\mathrm{tr}\left\lbrace\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace
setting above to zero and solving for $\sigma_{\epsilon}^2$ gives
 \sigma_{\epsilon}^2=\frac{1}{(T-1)n_y}\sum_{t=0}^{T-1}\mathrm{tr}\left\lbrace (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top \right\rbrace
A second derivative test shows that this is indeed the maximum. Finally differentiating with respec to $\sigma_d^2$ we get
 \frac{\partial \mathcal Q}{\partial \sigma_d^2}&= -(T-1)n_x\frac{1}{\sigma_d^2}+\frac{1}{\sigma_d^4}\mathrm{tr}\left\lbrace \boldsymbol\Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\frac{2\xi}{\sigma_d^4} \mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{\xi^2}{\sigma_d^4}\mathrm{tr} \left\lbrace\boldsymbol\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1} \right\rbrace  \nonumber \\
&-\frac{2}{\sigma_d^4}\boldsymbol\Xi_3\boldsymbol\theta +\frac{1}{\sigma_d^4}\boldsymbol\theta^\top \boldsymbol\Xi_4\boldsymbol\theta+\frac{2\xi}{\sigma_d^4} \boldsymbol\Xi_5 \boldsymbol\theta 
setting above to zero and solving for $\sigma_e^2$ we have
\sigma_d^2=\frac{1}{(T-1)n_x}\left[ \mathrm{tr}\left\lbrace \boldsymbol\Xi_0 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-
2\xi\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\xi^2\mathrm{tr} \left\lbrace\boldsymbol\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace -2\boldsymbol\Xi_3\boldsymbol\theta+\boldsymbol\theta^\top\boldsymbol \Xi_4\boldsymbol\theta+2\xi\boldsymbol\Xi_5 \boldsymbol\theta\right]  
In order to confirm calculated parameters are indeed maxima of the lower bound, the second derivative test is carried out for each parameter, giving
\begin{alignat}{2}%(number of &)+1 divided by 2 here we have (3+1)/2
 \frac{\partial \mathcal Q^2}{\partial^2\xi}&\biggr\rvert_{\xi=\hat{\xi}}&&=-\frac{2}{\sigma_d^2}\mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace \label{eq:SecondDerivativexi}\\
 \frac{\partial \mathcal Q^2}{\partial^2\boldsymbol\theta}&\biggr\rvert_{\boldsymbol\theta=\hat{\boldsymbol\theta}}&&=-\frac{2}{\sigma_d^2}\boldsymbol\Xi_4 \label{eq:SecondDerivativetheta} \\
 \frac{\partial \mathcal Q^2}{\partial^2\sigma_{\epsilon}^2}&\biggr\rvert_{\sigma_{\epsilon}=\hat{\sigma}_{\epsilon}}&&=-\frac{(T-1)n_y}{\hat{\sigma}_{\epsilon}^4} \label{eq:SecondDerivativenoise} \\
 \frac{\partial \mathcal Q^2}{\partial^2\sigma_{d}^2}&\biggr\rvert_{{\sigma}_{d}=\hat{\sigma}_{d}}&&= -\frac{(T-1)n_x}{\hat{\sigma}_{d}^4} \label{eq:SecondDerivativedisturbance}
\end{alignat}
$\boldsymbol\Xi_2$ is positive definite under the persistent excitation condition of state vectors, which is guaranteed by the disturbance, $\mathbf e$, and $\tilde{\boldsymbol\Sigma}_e^{-1}$ is positive definite by construction (Note that the inverse of a positive definite matrix is also positive definite \cite{Horn2005}), therefore we have \cite{Coope1994}
 0 < \mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace \le \mathrm{tr}\left\lbrace \boldsymbol\Xi_2\right\rbrace \mathrm{tr}\left\lbrace\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace,
hence \eqref{eq:SecondDerivativexi} is negative and $\hat{\xi}$ represents a maximum. Equation \eqref{eq:SecondDerivativetheta} is negative where $\boldsymbol\Xi_4$ is positive definite requiring $\mathbf q(\mathbf x_t)$ to be of  rank $n_{\theta}$ \cite{Golub1996}. \blue{don't know how to prove $q$ is of rank $n_{\theta}$} According to \eqref{eq:SecondDerivativetheta} and \eqref{eq:SecondDerivativedisturbance}, disturbance estimates clearly represent maxima of the $\mathcal{Q}$-function.
